---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->

<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->

<!-- Variables marked with double curly braces will be transformed beforehand: -->

<!-- `packagename` is extracted from the DESCRIPTION file -->

<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# **ColOpenData** <img src="man/figures/logo.svg" align="right" width="200"/>

<!-- badges: start -->

[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit) [![R-CMD-check](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml) [![Codecov test coverage](https://codecov.io/gh/epiverse-trace/ColOpenData/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/ColOpenData?branch=main) [![lifecycle-experimental](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-experimental.svg)](https://www.reconverse.org/lifecycle.html#experimental)

<!-- badges: end -->

**ColOpenData** is a package designed to access curated and wrangled Colombian demographic, geospatial, climate and population projections data, retrieved from various open Colombian data sources. The package addresses the challenge of scattered Colombian data across multiple web sources by providing functions that enable users to select and load desired datasets without the need for extensive data acquisition processes. Additionally, the tidy data structure offered for demographic and climate data facilitates analysis and visualization.

ColOpenData is developed at [Universidad de Los Andes](https://uniandes.edu.co/) as part of the [Epiverse-TRACE program](https://data.org/initiatives/epiverse/).

## Installation

You can install the CRAN version of ColOpenData with:

```{r eval=FALSE, include=TRUE}
install.packages("ColOpenData")
```

You can also install the development version of ColOpenData from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("epiverse-trace/ColOpenData")
```

## Quick Overview

**ColOpenData** contains data from two public data sources: Departamento Administrativo Nacional de Estadística (DANE), and Instituto de Hidrología, Meteorología y Estudios Ambientales (IDEAM). The available data is divided in four categories:

-   **Demographic:** Demographic and Socioeconomic data presents information from the National Population and Dwelling Census (CNPV) of 2018. The CNPV data corresponds to the most recent census available to date and the information is presented as an answer to three questions: How many are we?, Where are we? and How do we live?

-   **Geospatial:** This data is retrieved from the National Geostatistical Framework (MGN), which includes maps and a summarized version of the 2018 CNPV, aggregated to spatial geometries. The data is available at different aggregation levels including: Blocks, Urban and Rural Sections, Urban and Rural Sectors, Urban Areas, Municipalities and Departments.

-   **Climate:** Climate data is recovered from backup information provided by IDEAM, containing historical data from the first station in the country (January 1st 1920) until May 31st 2023. This backup includes temperature, precipitation, sunshine duration, wind direction, among others.

-   **Population projections:** Population Projections data contains the population projections and back projections from 1950 to 2070, considering the post COVID-19 update, which was calculated based on the results of 2018 CNPV.

Documentation and vignettes are available for the modules in the [user vignettes](https://epiverse-trace.github.io/ColOpenData/).

### Similar R Packages

Similar R packages are offered for international communities, allowing the user to download census, geospatial and climate data.

-   [cancensus](https://mountainmath.github.io/cancensus/): Canada
-   [censobr](https://ipeagit.github.io/censobr/): Brazil
-   [chilemapas](https://github.com/pachadotdev/chilemapas/): Chile
-   [geobr](https://ipeagit.github.io/geobr/) : Brazil
-   [georAr](https://github.com/PoliticaArgentina/geoAr): Argentina
-   [geouy](https://github.com/RichDeto/geouy): Uruguay
-   [tidycensus](https://walker-data.com/tidycensus/): US
-   [geofi](https://ropengov.github.io/geofi/): Finland
-   [climate](https://bczernecki.github.io/climate/)

### Disclaimer

ColOpenData stores and distributes open data originally obtained from IDEAM and 
DANE. It is important to note that the data retrieved from these sources has 
been processed, transformed, and adapted to meet the functional requirements of 
this package. These modifications may result in changes to the structure, 
format, or content of the original data, and as such, the data provided by this 
package does not represent the official datasets as maintained by IDEAM or DANE.
The development of this package is completely independent of these institutions, 
and neither IDEAM, DANE, nor any other Colombian government body endorses, 
sponsors, or is involved in the creation or maintenance of this package. 
Consequently, they bear no responsibility for the data as presented within this 
package. Furthermore, the authors of the ColOpenData assume no liability for any
outcomes resulting from the use or misuse of the data provided. The user is 
solely responsible for how they utilize the data, and any conclusions drawn, 
analyses performed, or actions taken based on this data are the user's 
responsibility. This package and its authors make no claims to represent or act 
on behalf of IDEAM, DANE, or any other official entity.

### Lifecycle

This package is currently *experimental*, as defined by the [RECON software lifecycle](https://www.reconverse.org/lifecycle.html). Therefore, this is a functional draft and can be tested outside of the development team. However, it still may change over time.

### Contributions

Contributions are welcome via [pull requests](https://github.com/epiverse-trace/ColOpenData/pulls).

### Code of Conduct

Please note that the ColOpenData project is released with a [Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md). By contributing to this project, you agree to abide by its terms.

### Funding

This work is part of the TRACE-LAC research project funded by the International Research Centre (IDRC) Ottawa, Canada.[109848-001-].
