---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ColOpenData <img src="man/figures/logo.svg" align="right" width="120" />

<!-- badges: start -->

[![License:
MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml)
[![Codecov test
coverage](https://codecov.io/gh/epiverse-trace/ColOpenData/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/ColOpenData?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept)
<!-- badges: end -->

ColOpenData is a package that acquires, standardizes, and wrangles Colombian
socioeconomic, climate and land cover data. It solves the problem of
Colombian data being issued in different web pages and formats by using functions that
allow the user to select the desired database and download it without having to do the
exhausting acquisition process. Also, it allows these datasets to be merged so multiple
operations, such as calculating statistics, can be made, taking into account information
from different datasets.

<!-- This sentence is optional and can be removed -->
ColOpenData is developed at [Universidad de Los
Andes](https://uniandes.edu.co/) as part of the [Epiverse-TRACE
program](https://data.org/initiatives/epiverse/).

## Installation

You can install the development version of ColOpenData from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("epiverse-trace/ColOpenData")
```

## Quick start
This example shows how to retrieve census data at a department level in Colombia
including the administrative divisions and spatial data, to later use it to 
visualize Dengue cases during 2018. We will first load the needed packages.
```{r load_packages, message= FALSE}
library(ColOpenData)
library(sf)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
```

We will be using the `MGNCNPV_DPTO_2018` dataset, which contains the National 
Geo statistical  Framework (MGN) and the National Population and Living Census 
(CNPV) at department level for 2018.

To load the dataset we can use the `download` function as follows.
```{r download_data}
census <- download("MGNCNPV_DPTO_2018")
str(census)
```

Once downloaded, we can subset the data by regions or columns of interest. In this example we only need the departments codes and geometries.
```{r filter_data}
col <- census %>%
  select(DPTO_CCDGO, geometry)
```

Dengue cases during 2018 can be loaded from the example dataset `dengue_2018`,
which was previously obtained using the package [sivirep](https://epiverse-trace.github.io/sivirep/).
```{r load_data}
dengue_cases <- dengue_2018
str(dengue_cases)
```

To merge the datasets we will use the department's codes.
```{r}
dengue_col <- merge(
  x = col,
  y = dengue_2018,
  by = "DPTO_CCDGO"
)
```

Finally, we can visualize the cases in a map.
```{r plot_data}
ggplot(data = dengue_col) +
  geom_sf(mapping = aes(fill = CASES)) +
  scale_fill_gradientn(
    colours = brewer.pal(7, "YlOrRd"),
    name = element_blank()
  ) +
  ggtitle("Dengue Cases in Colombia during 2018") +
  theme_bw()
```


### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/ColOpenData/pulls).

### Code of Conduct

Please note that the ColOpenData project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
