---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file. -->
<!-- The code to render this README is stored in .github/workflows/render-readme.yaml -->
<!-- Variables marked with double curly braces will be transformed beforehand: -->
<!-- `packagename` is extracted from the DESCRIPTION file -->
<!-- `gh_repo` is extracted via a special environment variable in GitHub Actions -->


```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# ColOpenData <img src="man/figures/logo.svg" align="right" width="120" />

<!-- badges: start -->

[![License:
MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://opensource.org/license/mit/)
[![R-CMD-check](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/epiverse-trace/ColOpenData/actions/workflows/R-CMD-check.yaml)
[![Codecov test
coverage](https://codecov.io/gh/epiverse-trace/ColOpenData/branch/main/graph/badge.svg)](https://app.codecov.io/gh/epiverse-trace/ColOpenData?branch=main)
[![lifecycle-concept](https://raw.githubusercontent.com/reconverse/reconverse.github.io/master/images/badge-concept.svg)](https://www.reconverse.org/lifecycle.html#concept)
<!-- badges: end -->

ColOpenData is a package that acquires, standardizes, and wrangles Colombian
socioeconomic, climate and land cover data. It solves the problem of
Colombian data being issued in different web pages and formats by using functions that
allow the user to select the desired database and download it without having to do the
exhausting acquisition process. Also, it allows these datasets to be merged so multiple
operations, such as calculating statistics, can be made, taking into account information
from different datasets.

<!-- This sentence is optional and can be removed -->
ColOpenData is developed at [Universidad de Los
Andes](https://uniandes.edu.co/) as part of the [Epiverse-TRACE
program](https://data.org/initiatives/epiverse/).

## Installation

You can install the development version of ColOpenData from
[GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("epiverse-trace/ColOpenData")
```

## Quick start
This example shows how to retrieve census data at a department level in Colombia
including the administrative divisions and spatial data. We will be using the
`MGNCNPV01` dataset, that contains the National Geo statistical Framework (MGN)
and the National Population and Living Census (CNPV), and subset a certain
region to calculate the population density on each department of this region.

```{r load_packages, message= FALSE}
library(ColOpenData)
library(sf)
library(dplyr)
library(ggplot2)
library(RColorBrewer)
```

To download the dataset we can use the `download` function as follows
```{r download_data}
census <- download("MGNCNPV01")
```

Once downloaded, we can filter the data to subset only the regions of interest
using the `filter_mgn_cnpv` function. We will be selecting the departments in 
the Amazonian region using the [DIVIPOLA codes](https://www.datos.gov.co/widgets/gdxc-w37w) 
of each, and to calculate the population density we'll select the
population and area.

```{r filter_data}
amazonia <- census %>%
  filter_mgn_cnpv(
    codes = c(91, 18, 19, 94, 95, 50, 52, 86, 97, 99),
    columns = c("DPTO_CNMBR", "STP27_PERS", "AREA"),
    include_geometry = TRUE
  )
```

As the area is stated in $m^2$ we need to transform it to $km^2$ and calculate 
the density in each department.
```{r mutate_data}
amazonia <- amazonia %>%
  mutate(
    AREA = AREA / 1000000,
    DENSITY = STP27_PERS / AREA
  )
```

Finally, to visualize the data we will plot a map with the density.
```{r plot_data}
ggplot(data = amazonia) +
  geom_sf(mapping = aes(fill = DENSITY)) +
  scale_fill_gradientn(
    colours = brewer.pal(7, "YlOrRd"),
    name = expression("Density [" * People / km^2 * "]")
  ) +
  ggtitle("Population density by square kilometer") +
  theme_bw()
```


### Lifecycle

This package is currently a *concept*, as defined by the [RECON software
lifecycle](https://www.reconverse.org/lifecycle.html). This means that essential
features and mechanisms are still being developed, and the package is not ready
for use outside of the development team.


### Contributions

Contributions are welcome via [pull requests](https://github.com/ColOpenData/pulls).

### Code of Conduct

Please note that the ColOpenData project is released with a 
[Contributor Code of Conduct](https://github.com/epiverse-trace/.github/blob/main/CODE_OF_CONDUCT.md).
By contributing to this project, you agree to abide by its terms.
