---
title: "Download climate data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download climate data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**ColOpenData** can be used to access open climate data from Colombia. This
climate data is retrieved from the Institute of Hydrology, Meteorology and 
Environmental Studies [(IDEAM)]( http://www.ideam.gov.co/). The climate module
allows you to consult climate data for any Region of Interest (ROI) inside the
country and aggregate for different frequencies like day, week, month or year.

The available information from IDEAM can be accessed using specific internal 
tags as follows:

```{r IDEAM table, echo = FALSE}
tags <- c(
  "TSSM_CON", "THSM_CON", "TMN_CON", "TMX_CON", "TSTG_CON", "HR_CAL",
  "HRHG_CON", "TV_CAL", "TPR_CAL", "PTPM_CON", "PTPG_CON", "EVTE_CON",
  "FA_CON", "NB_CON", "RCAM_CON", "BSHG_CON", "VVAG_CON", "DVAG_CON",
  "VVMXAG_CON", "DVMXAG_CON"
)
variable <- c(
  "Dry-bulb Temperature", "Wet-bulb Temperature",
  "Minimum Temperature", "Maximum Temperature",
  "Dry-bulb Temperature (Termograph)", "Relative Humidity",
  "Relative Humidity (Hydrograph)", "Vapour Pressure", "Dew Point",
  "Precipitation (Daily)", "Precipitation (Hourly)", "Evaporation",
  "Atmospheric Phenomenon", "Cloudiness", "Wind Trajectory",
  "Solar Luminosity", "Wind Speed", "Wind Direction",
  "Maximum Wind Speed", "Maximum Wind Direction"
)

IDEAM_tags <- data.frame(
  Tags = tags, Variable = variable,
  stringsAsFactors = FALSE
)
knitr::kable(IDEAM_tags)
```

Each variable's observation is subject to to the availability of stations in the
ROI and the stations' status (active, maintenance or suspended). 

For this example we will retrieve data for the city of Manizales in Colombia. 
We will download Dry-bulb temperature from 2013 to 2016, to observe the
increase in the temperature during 2015 and 2016 due to the impact of El 
Nino (ENSO).

We will first load the necessary libraries.

```{r library imports, results = "hide", warning = FALSE, message = FALSE}
library(ColOpenData)
library(sf)
library(leaflet)
library(ggplot2)
```

## Retrieving climate data for a Region of Interest (ROI)

To retrieve climate data for any ROI in the country, we can use the
function `download_climate_geom`, which downloads data from climate stations 
contained inside a given geometry from a ROI. The function has the following
parameters:

- `geometry`: `sf` object containing the geometry for a given ROI. This geometry
can be either a POLYGON or MULTIPOLYGON.(see [sf](https://r-spatial.github.io/sf/articles/sf1.html) library for further details)
- `start_date`: Character with the first date to consult in format "YYYY-MM_DD"
- `end_date`: Character with the last date to consult in format "YYYY-MM_DD"
- `frequency`: Character with the aggregation frequency. Can be "day", "week",
"month" or"year"
- `tags`: Character containing tags to consult.
- `group`: Boolean for data aggregation, if FALSE, returns the data from each
individual station in ROI.(Default = FALSE)

For this example, we will need to create a spatial polygon around the city of
Manizales and use that as our ROI to retrieve the climate data.

```{r polygon creation}
# Define coordinates (Bounding box around Manizales)
lat <- c(5.166278, 5.166278, 4.982247, 4.982247, 5.166278)
lon <- c(-75.678072, -75.327859, -75.327859, -75.678072, -75.678072)

# Use sf to create a polygon and turn it into a geometry
polygon <- st_polygon(x = list(cbind(lon, lat))) %>% st_sfc()

# Turn geometry into sf object
roi <- st_as_sf(polygon)

# plot polygon for reference
leaflet(roi) %>%
  addProviderTiles("OpenStreetMap") %>%
  addPolygons(
    stroke = TRUE,
    weight = 2,
    color = "#2e6930",
    fillColor = "#2e6930",
    opacity = 0.6
  )
```

With our new ROI, we can make a first exploration to check if there are any 
stations inside the ROI, using the function `stations_in_roi`
```{r stations in roi, results = "hide", warning = FALSE, message = FALSE}
stations <- stations_in_roi(roi)
```


```{r print stations}
print(stations)
```
We can see that in the region there are 129 stations with different categories,
which means not all of them hold record for the same variables. Also, a lot of
them were suspended, which means they are no longer taking any extra 
observations, however, we still can access the historical data.

Now that we  have confirmed that there are meteorological stations in the area,
we will access the Dry-bulb temperature from 2013 to 2016. To do this we will
use `download_climate`. The Dry-bulb temperature is under the tag "TSSM_CON". 

```{r download climate data, results = "hide", warning = FALSE, message = FALSE}
db_temperature_roi <- download_climate_geom(
  geometry = roi,
  start_date = "2013-01-01",
  end_date = "2016-12-31",
  frequency = "day",
  tags = "TSSM_CON",
  group = TRUE
)
```

With the recorded data we can plot the time series, for the four years.
```{r plot temperatures}
ggplot(data = db_temperature_roi) +
  geom_line(aes(x = dates, y = TSSM_CON)) +
  ggtitle("Dry-bulb Temperature") +
  xlab("Date") +
  ylab("Dry-bulb temperature [C]") +
  theme_bw()
```

## Retrieving climate data for municipality

To make the download process easier, and avoid the creation of already known
geometries like municipalities, **ColOpenData** offers a different function, to
download data using the municipalities DIVIPOLA code. This codification is 
standardized for the whole country, and contains a department code (first two 
digits) and the municipality code (last three digits), for a final five digit
string. The codes for each municipality and department can be consulted in the
following [table](https://www.datos.gov.co/widgets/gdxc-w37w). To
replicate our previous example we only need the code for Manizales, which is
"17001".

The function `download_climate_mpio` will require almost the same arguments 
than `download_climate`, :

- `code`: Character with the DIVIPOLA code for the municipality.
- `start_date`: Character with the first date to consult in format "YYYY-MM_DD".
- `end_date`: Character with the last date to consult in format "YYYY-MM_DD".
- `frequency`: Character with the aggregation frequency. Can be "day", "week",
"month" or"year".
- `tags`: Character containing tags to consult.
- `group`: Boolean for data aggregation, if FALSE, returns the data from each
individual station in the municipality.(Default = FALSE).

```{r download climate mpio, results = "hide", warning = FALSE, message = FALSE}
db_temperature_mpio <- download_climate_mpio(
  code = "17001",
  start_date = "2013-01-01",
  end_date = "2016-12-31",
  frequency = "day",
  tags = "TSSM_CON",
  group = TRUE
)
```
```{r}
ggplot(data = db_temperature_mpio) +
  geom_line(aes(x = dates, y = TSSM_CON)) +
  ggtitle("Dry-bulb Temperature") +
  xlab("Date") +
  ylab("Dry-bulb temperature [C]") +
  theme_bw()
```


