---
title: "A Deep Dive into Colombian Demographics Using ColOpenData"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A Deep Dive into Colombian Demographics Using ColOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {style="text-align: justify;"}
**ColOpenData** can be used to access open demographic data from Colombia. This demographic data is retrieved from the National Administrative Department of Statistics ([DANE](https://www.dane.gov.co/index.php/estadisticas-por-tema)). The demographic module allows you to consult demographic data from the National Population and Dwelling Census (CNPV) of 2018 and Population Projections.

The available CNPV information is divided in four categories: households, persons demographic, persons social and dwellings. The population projections information presents data from 1950 to 2070 for a national level, from 1985 to 2050 for a departmental level and from 1985 to 2035 for a municipal level. All data documentation can be accessed as explained at [Documentation and Dictionaries](https://epiverse-trace.github.io/ColOpenData/articles/documentation_and_dictionaries.html).

In this vignette you will learn:

1. How to download demographic data using **ColOpenData**
2. How to filter, group, mutate and aggregate demographic data including census and population projections
3. How to visualize data using **ggplot2**

As the goal of this vignette is to show some examples on how to use the data, we will load some specific libraries but that does not mean they are required to use the data in all cases.

In order to access its documentation we need to use the function `list_datasets` and indicate as a parameter the module we are interested in. It is important to take a good look at this to have a clearer understanding of what we count with, before just throwing ourselves to work with the data. Now, we should start by loading all necessary libraries. 
:::

```{r libraries, message=FALSE, warning=FALSE, results="hide"}
library(ColOpenData)
library(dplyr)
library(ggplot2)
```

::: {style="text-align: justify;"}
**Disclaimer: all data is loaded to the environment in the user's R session, but is not downloaded to user's computer.**
:::

## Initial Exploration: Basic Data Handling with ColOpenData

### Documentation access

::: {style="text-align: justify;"}
First, we have to access the demographic documentation, to check available datasets.
:::

```{r documentation, echo =TRUE}
datasets_dem <- list_datasets("demographic")

head(datasets_dem)
```
::: {style="text-align: justify;"}
After checking the documentation, we can load the data we want to work with. To do this, we will use the `download_demographic` function that takes by parameter the dataset name, presented in the documentation. For this first example we will focus on a CNPV dataset.
:::

### Data load

```{r data load, echo=TRUE}
public_services_d <- download_demographic("DANE_CNPVV_2018_8VD")

head(public_services_d)
```

::: {style="text-align: justify;"}
As it can be seen above, [public_services_d]{.underline} presents information regarding availability of public services in the country at the department level. Now, with this data we could, for example, find the proportion of dwellings that have access to a water supply system (WSS) by department and plot it.
:::

### Data filter and plot

First we will subset the data so it presents the information regarding the WSS by department.

```{r wss}
wss <- public_services_d %>%
  filter(
    area == "total_departamental",
    servicio_publico == "acueducto"
  ) %>%
  select(departamento, disponible, total)
```

With the subset, we can calculate the total counts by department.

```{r counts wss}
total_counts <- wss %>%
  group_by(departamento) %>%
  summarise(total_all = sum(total)) %>%
  ungroup()
```

Then, we can calculate the proportions of "yes" ("si") by department.

```{r proportions wss}
proportions_wss <- wss %>%
  filter(disponible == "si") %>%
  left_join(total_counts, by = "departamento") %>%
  mutate(proportion_si = total / total_all)
```

For plotting purposes, we will change the name of "San Andrés", since the complete name is too long.

```{r san andres}
proportions_wss[28, "departamento"] <- "SAPSC"
```

Finally, we can plot the results

```{r plot wss}
ggplot(proportions_wss, aes(
  x = reorder(departamento, -proportion_si),
  y = proportion_si
)) +
  geom_bar(stat = "identity", fill = "#10bed2", color = "black", width = 0.6) +
  labs(
    title = "Proportion of dwellings with access to WSS by department",
    x = "Department",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

## Data Integration: Combining Datasets for Enhanced Analysis

::: {style="text-align: justify;"}
Also, with the CNPV data, we could merge data from the different categories and compare information. For example, we would download a dataset from the persons demographic category and another one from persons social. The first one presents information regarding population's literacy and the second one regarding its marital status. The following datasets might take a while to download.
:::

### Data load

```{r more data load, include=TRUE, echo=TRUE}
literacy <- download_demographic("DANE_CNPVPD_2018_16PM")
marital_status <- download_demographic("DANE_CNPVPS_2018_7PM")
```
::: {style="text-align: justify;"}
For this example, we will focus on a specific municipality, examining the total area and specifically concentrating on the age group between 15 and 24 years old. Since we want to compare both datasets simultaneously, it's important to first ensure they have the same population count under identical conditions. Most columns in the demographic datasets contain a [total]{.underline} value, which aggregates data across all conditions. For instance, the "sex" column includes "male", "female", and ["total"]{.underline}, representing both genders combined. Similarly, the "marital status" column's ['total']{.underline} value encompasses the entire population, irrespective of their marital status.
:::

### Data check

::: {style="text-align: justify;"}
Since the age group in literacy is complete, we can just filter using the desired age range and the total counts for other variables, including area, sex and ability to read.
:::

```{r literacy}
lit <- literacy %>%
  filter(
    municipio == "Villapinzón",
    grupo_de_edad == "entre_15_y_24_anos",
    area == "total",
    sabe_leer_y_escribir == "total",
    sexo == "total"
  ) %>%
  select(sexo, total)

head(lit)
```

::: {style="text-align: justify;"}
We will now filter marital status for people between 15 and 24 years. This dataset age groups are divided from 15 to 19 and from 20 to 24 years old, so we will need to merge the two groups. To do this, we can group by a common variable, like sex and add as follows:
:::

```{r}
ms_15_24 <- marital_status %>%
  filter(
    municipio == "Villapinzón",
    area == "total",
    grupo_de_edad %in% c("15_a_19", "20_a_24"),
    estado_civil == "total",
    sexo == "total"
  ) %>%
  group_by(sexo) %>%
  summarise(total = sum(total)) %>%
  select(sexo, total)

head(ms_15_24)
```

::: {style="text-align: justify;"}
Now that we know that both data sets contain the same counts, we can start to look at the specific information we want. For the literacy data set, we would focus on the people that do not know how to read neither write.
:::

### Data filter and plot

::: {style="text-align: justify;"}
We can filter the literacy data. For this purpose we will filter the column "sabe_leer_y_escribir" for the value "no". We will also create a column containing the proportion of people that cannot read compared to the total population within the age range.
:::

```{r}
lit_filt <- literacy %>%
  filter(
    municipio == "Villapinzón",
    sabe_leer_y_escribir == "no",
    grupo_de_edad == "entre_15_y_24_anos",
    area == "total"
  ) %>%
  mutate(
    estado = "no_sabe_leer_y_escribir",
    proporcion = total / lit$total
  ) %>%
  select(estado, sexo, proporcion)
```

::: {style="text-align: justify;"}
To filter the marital status dataset we will filter by the two age groups again. We will group by marital status and sex, sum by the total counts and create a column containing the proportion of each marital status compared to the population within the age group.
:::

```{r}
ms_15_24_filt <- marital_status %>%
  filter(
    municipio == "Villapinzón",
    area == "total",
    grupo_de_edad %in% c("15_a_19", "20_a_24")
  ) %>%
  group_by(estado_civil, sexo) %>%
  summarise(total = sum(total)) %>%
  mutate(proporcion = total / ms_15_24$total) %>%
  select(estado_civil, sexo, proporcion) %>%
  rename(estado = estado_civil)
```

Finally, we can merge the two datasets and plot the outcome.

```{r}
merged_df <- bind_rows(lit_filt, ms_15_24_filt) %>%
  filter(
    sexo != "total",
    estado != "total"
  )

ggplot(merged_df, aes(x = estado, y = proporcion, fill = sexo)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c(mujer = "#f04a4c", hombre = "#aec800"),
    labels = c("Female", "Male")
  ) +
  labs(
    title = "Proportion of people in Villapinzón by marital status and
    iliteracy by sex",
    x = "Status",
    y = "Proportion",
    fill = "Sex"
  ) +
  theme_bw() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

## Forecasting the Future: Analyzing Population Projections

::: {style="text-align: justify;"}
Now, as a third example, we will take a look at a population projections dataset, presenting projections and back projections of national population by area, sex and age for the period from 1950 to 2070. We will observe the expected female population under 99 by age brackets for 2034.
:::

We will first download the data.

```{r download data}
asen <- download_demographic("DANE_PP_1950_2070_ASEN")
```
We will filter the downloaded data for the year 2034 and ages under 99.

```{r filtered projections}
female_2034 <- asen %>%
  filter(
    area == "total",
    ano == 2034,
    sexo == "mujer",
    edad != "100_y_mas"
  ) %>%
  mutate(edad = as.numeric(edad))
```

Age groups will be defined by breaks and included in the original dataset.

```{r age groups}
age_groups <- cut(female_2034$edad,
  breaks = c(-1, 2, 12, 19, 29, 39, 49, 59, 69, 79, 89, 99),
  labels = c(
    "0-2", "3-12", "13-19", "20-29", "30-39", "40-49",
    "50-59", "60-69", "70-79", "80-89", "90-99"
  )
)
female_groups <- female_2034 %>%
  mutate(age_group = age_groups) %>%
  group_by(age_group) %>%
  summarise(total_sum = sum(total))
```

Finally, we can plot the output.

```{r plot population}
ggplot(female_groups, aes(
  x = age_group,
  y = total_sum
)) +
  geom_bar(stat = "identity", fill = "#f04a4c", color = "black", width = 0.6) +
  labs(
    title = "Female population counts in Colombia by age group for 2034",
    x = "Age group",
    y = "Female population"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

