---
title: "A Deep Dive into Colombian Demographics Using ColOpenData"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{A Deep Dive into Colombian Demographics Using ColOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {style="text-align: justify;"}
**ColOpenData** can be used to access open demographic data from Colombia. This demographic data is retrieved from the National Administrative Department of Statistics ([DANE](https://www.dane.gov.co/index.php/estadisticas-por-tema)). The demographic module allows you to consult demographic data from the National Population and Dwelling Census (CNPV) of 2018 and Population Projections.

The available CNPV information is divided in four categories: households, persons demographic, persons social and dwellings. The population projections information presents data from 1950 to 2070 for a national level, from 1985 to 2050 for a departmental level and from 1985 to 2035 for a municipal level. All data documentation can be accessed as explained at [Documentation and Dictionaries](https://epiverse-trace.github.io/ColOpenData/articles/documentation_and_dictionaries.html).

In order to access its documentation we need to use the function `list_datasets` and indicate as a parameter the module we are interested in. It is important to take a good look at this to have a clearer understanding of what we count with, before just throwing ourselves to work with the data. Now, we should start by loading all necessary libraries.

In this vignette you will learn: 1. How to download demographic data using **ColOpenData** 2. How to filter, group, mutate and aggregate demographic data including census and population projections 3. How to visualize data using **ggplot2**

As the goal of this vignette is to show some examples on how to use the data, we will load some specific libraries but that does not mean they are required to use the data in all cases.
:::

```{r libraries, message=FALSE, warning=FALSE, results="hide"}
library(ColOpenData)
library(dplyr)
library(ggplot2)
```

::: {style="text-align: justify;"}
**Disclaimer: all data is loaded to the environment in the user's R session, but is not downloaded to user's computer.**
:::

## Initial Exploration: Basic Data Handling with ColOpenData

### Documentation access

::: {style="text-align: justify;"}
First, we have to access the demographic documentation, to check available datasets.

```{r documentation, echo =TRUE}
datasets_dem <- list_datasets("demographic")
head(datasets_dem)
```

After checking the documentation, we can load the data we want to work with. To do this, we will use the `download_demographic` function that takes by parameter the data name, presented in the documentation. For this first example we will focus on a CNPV dataset.
:::

### Data load

```{r data load, echo=TRUE}
# Load a demographic dataset
public_services_d <- download_demographic("DANE_CNPVV_2018_8VD")
head(public_services_d)
```

::: {style="text-align: justify;"}
As it can be seen above, [public_services_d]{.underline} presents information regarding availability of public services in the country in 2018, at the department level. Now, with this data we could, for example, find the proportion of dwellings that have access to a water supply system (WSS) by department and plot it.
:::

### Data filter and plot

```{r data filter and plot, echo=TRUE, fig.width=7, fig.height=3}
# Subset data so it presents the information regarding the WSS, by department
wss <- public_services_d %>%
  subset(area == "total_departamental" & servicio_publico == "acueducto") %>%
  select(departamento, disponible, total)

# Calculate the total counts for each department (for all situations)
total_counts <- wss %>%
  group_by(departamento) %>%
  dplyr::summarise(total_all = sum(total)) %>%
  ungroup()

# Calculate the proportion of counts of "si" by department
proportions_wss <- wss %>%
  filter(disponible == "si") %>%
  left_join(total_counts, by = "departamento") %>%
  mutate(proportion_si = total / total_all)

# Change the value of cell San Andres
proportions_wss[28, "departamento"] <- "SAPSC"

# Plot proportions_wss of dwellings with access to WSS by deparment
ggplot(proportions_wss, aes(
  x = reorder(departamento, -proportion_si),
  y = proportion_si
)) +
  geom_bar(stat = "identity", fill = "#10bed2", color = "black", width = 0.6) +
  labs(
    title = "Proportion of dwellings with access to WSS, by department",
    x = "Department",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

## Data Integration: Combining Datasets for Enhanced Analysis

::: {style="text-align: justify;"}
Also, with the CNPV data, we could merge data from the different categories and compare information. For example, we would download a dataset from the persons demographic category and another one from persons social. The first one presents information regarding population's literacy and the second one regarding its marital status.
:::

### Data load

```{r more data load, include=TRUE, echo=TRUE}
# Load data
message("Loading data, this can take a while")
literacy <- download_demographic("DANE_CNPVPD_2018_16PM")
marital_status <- download_demographic("DANE_CNPVPS_2018_7PM")
```

::: {style="text-align: justify;"}
For this example, we will focus on a specific municipality, examining the total area and specifically concentrating on the age group between 15 and 24 years old. Since we want to compare both datasets simultaneously, it's important to first ensure they have the same population count under identical conditions. Most columns in the demographic datasets contain a [total]{.underline} value, which aggregates data across all conditions. For instance, the 'sex' column includes 'male', 'female', and [total]{.underline}, representing both genders combined. Similarly, the 'marital status' column's [total]{.underline} value encompasses the entire population, irrespective of their marital status.
:::

### Data check

```{r population counts, message=FALSE, warning=FALSE, results="hide"}
# Literacy data filter to have a population's count under specific conditions
lit <- literacy %>%
  subset(municipio == "Villapinzón" & grupo_de_edad == "entre_15_y_24_anos" &
    area == "total" & sabe_leer_y_escribir == "total" & sexo == "total") %>%
  select(municipio, area, grupo_de_edad, sexo, total)

# Marital status data filter to find population's count under certain conditions
ms_15_19 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "15_a_19" & estado_civil == "total" &
    sexo == "total") %>%
  select(grupo_de_edad, estado_civil, sexo, total)

ms_20_24 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "20_a_24" & estado_civil == "total" &
    sexo == "total") %>%
  select(grupo_de_edad, estado_civil, sexo, total)

ms <- rbind(ms_15_19, ms_20_24)

# Sum of counts of both age groups
ms_count <- ms %>%
  group_by(estado_civil, sexo) %>%
  dplyr::summarise(total = sum(total)) %>%
  ungroup()
count_total <- toString(ms_count$total)

# New row with summed information
new_row <- data.frame(
  grupo_de_edad = "15_a_24", estado_civil = "total",
  sexo = "total", total = count_total,
  stringsAsFactors = FALSE
)
# Final dataframe with counts for each group independently and together, with
# rows names modification

ms_f <- rbind(ms, new_row)
rownames(ms_f) <- NULL
head(ms_f)
```

::: {style="text-align: justify;"}
Now that we know that both data sets contain the same counts, we can start to look at the specific information we want. For the literacy data set, we would focus on the people that do not know how to read neither write.
:::

### Data filter and plot

```{r other data plot, echo=TRUE, fig.width=10, fig.height=3}
# Filter of literacy data to get only the counts for specific conditions
lit_filt <- literacy %>%
  subset(municipio == "Villapinzón" & sabe_leer_y_escribir == "no" &
    grupo_de_edad == "entre_15_y_24_anos" & area == "total") %>%
  select(sabe_leer_y_escribir, grupo_de_edad, sexo, total) %>%
  mutate(estado = "no_sabe_leer_y_escribir") %>%
  select(-sabe_leer_y_escribir) %>%
  mutate(proportion = total / lit$total)

# Filter of marital status data to get only the counts for specific conditions
ms_filt15 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "15_a_19") %>%
  select(grupo_de_edad, estado_civil, sexo, total)

# Filter of marital status data to get only the counts for specific conditions
ms_filt20 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "20_a_24") %>%
  select(grupo_de_edad, estado_civil, sexo, total)

# Merge of both marital status dataframes
ms_filt <- rbind(ms_filt15, ms_filt20)

# Data frame with marital status info and proportions based on columns
ms_final <- ms_filt %>%
  group_by(estado_civil, sexo) %>%
  dplyr::summarise(total = sum(total)) %>%
  ungroup() %>%
  mutate(grupo_de_edad = "entre_15_y_24_anos") %>%
  rename(estado = estado_civil) %>%
  mutate(proportion = total / lit$total)

# Merged marital status and literacy final df, with subsets to a better graph
merged_df <- bind_rows(lit_filt, ms_final) %>%
  subset(sexo != "total") %>%
  subset(estado != "total") %>%
  subset(estado != "soltero_a")

# Plot
ggplot(merged_df, aes(x = estado, y = proportion, fill = sexo)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(
    values = c(mujer = "#f04a4c", hombre = "#aec800"),
    labels = c("Female", "Male")
  ) +
  labs(
    title = "Proportion of people in Villapinzón by marital status and
    iliteracy by sex",
    x = "Status",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

## Forecasting the Future: Analyzing Population Projections

::: {style="text-align: justify;"}
Now, as a third example, we will take a look at a population projections dataset presenting projections and back projections of national population by area, sex and age for the period from 1950 to 2070.

First, we will observe the variation on the female and male population within the years, covering the period from 2024 to 2034.
:::

### Data load, filter and plot

```{r pop projections first example, echo=TRUE}
# Download data
asen <- download_demographic("DANE_PP_1950_2070_ASEN")

# Subset data
male <- asen %>%
  subset(area == "total" & sexo == "total_hombres" & ano > 2023 & ano < 2035)

female <- asen %>%
  subset(area == "total" & sexo == "total_mujeres" & ano > 2023 & ano < 2035)

mandf <- bind_rows(male, female)

# Plot of population in a national level by year and sex
ggplot(mandf, aes(x = ano, y = total, fill = sexo)) +
  geom_bar(stat = "identity", position = "dodge", width = 0.6) +
  scale_fill_manual(
    values = c(total_mujeres = "#f04a4c", total_hombres = "#aec800"),
    labels = c("Female", "Male")
  ) +
  labs(
    title = "Population in Colombia from 2024 to 2034 by sex",
    x = "Year",
    y = "Population",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  ) +
  scale_x_continuous(breaks = unique(mandf$ano))
```

::: {style="text-align: justify;"}
In the graph presented above, we observe that the male population slightly exceeds the female population for the selected period. As a second step in our analysis of the dataset, we will focus on the male population by age in the year with the highest male population value from 2024 to 2034, which is the year 2034.
:::

### Data load, filter and plot

```{r pop projections second example, echo=TRUE}
# Subset data
male_2034_99 <- asen %>%
  subset(area == "total" & ano == 2034 & sexo == "hombre" &
    edad != "100_y_mas") %>%
  mutate(edad = as.numeric(edad))

male_2034_100 <- asen %>%
  subset(area == "total" & ano == 2034 & sexo == "hombre" &
    edad == "100_y_mas") %>%
  rename(age_group = edad) %>%
  rename(total_sum = total) %>%
  select(age_group, total_sum)

# Define age groups
age_groups <- cut(male_2034_99$edad,
  breaks = c(-1, 2, 12, 19, 29, 39, 49, 59, 69, 79, 89, 99),
  labels = c(
    "0-2", "3-12", "13-19", "20-29", "30-39", "40-49",
    "50-59", "60-69", "70-79", "80-89", "90-99"
  )
)

# Add age groups to the data frame
male_agegroups <- mutate(male_2034_99, age_group = age_groups)

# Calculate new totals for age groups
summarized_data <- male_agegroups %>%
  group_by(age_group) %>%
  summarize(total_sum = sum(total))

# Bind dataframes
merged <- bind_rows(summarized_data, male_2034_100)

# Fix the order of ages appearance in the plot
merged$age_group <- factor(merged$age_group, levels = unique(merged$age_group))

# Plot of male population in a national level by age group in 2052
ggplot(merged, aes(
  x = age_group,
  y = total_sum
)) +
  geom_bar(stat = "identity", fill = "#f04a4c", color = "black", width = 0.6) +
  labs(
    title = "Male population counts in Colombia by age group in 2034",
    x = "Age group",
    y = "Male population"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```
