---
title: "A Deep Dive into Colombian Demographics Using ColOpenData"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{A Deep Dive into Colombian Demographics Using ColOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{css, include = FALSE}
/* CSS rules for adjusting plot size */
.my-plot {
    width: 100%; /* Adjust the width as needed */
    height: auto; /* Allow the height to adjust proportionally */
    margin: auto; /* Center the plot horizontally */
}

```

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {style="text-align: justify;"}

**ColOpenData** can be used to access open demographic data from Colombia. This demographic data is retrieved from the National Administrative Department of Statistics ([DANE](https://www.dane.gov.co/index.php/estadisticas-por-tema), for its acronym in Spanish). The demographic module allows you to consult demographic data, more specifically, National Population and Dwelling Census (CNPV) of 2018.

The available CNPV information is divided in four categories: households, persons demographic, persons social and dwellings. All data documentation can be accessed as explained at [Documentation and Dictionaries](https://epiverse-trace.github.io/ColOpenData/articles/documentation_and_dictionaries.html). Just as a reminder, the documentation presents information regarding all the three modules of the package, but here we are just interested in the demographic one. In order to access its documentation we need to use the function `list_datasets` and by parameter indicate the module we are interested in. It is important to take a good look at this to have a clearer understanding of what we count with, before just throwing ourselves to work with the data. Now, we should start by loading all necessary libraries. As the goal of this vignette is to show some examples on how to use the data, we will load some specific libraries but that does not mean they are required to use the data in all cases.
:::

## Initial Exploration: Basic Data Handling with ColOpenData

```{r libraries, message=FALSE, warning=FALSE, results="hide"}
library(ColOpenData)
library(dplyr)
library(ggplot2)
```

::: {style="text-align: justify;"}
Then, we can access to the demographic documentation.

```{r documentation, echo =TRUE}
datasets_dem <- list_datasets("demographic")
head(datasets_dem)
```

After checking the documentation, we can download the data we want to work with. To do this, we will use the `download_demographic` function that takes by parameter the data name, presented in the documentation.
:::

```{r data download, echo=TRUE}
# Download a demographic dataset
public_services_d <- download_demographic("DANE_CNPVV_2018_8VD")
head(public_services_d)
```

::: {style="text-align: justify;"}
As it can be seen above, [public_services_d]{.underline} presents information regarding availability of public services in the country in 2018, at the department level. Now, with this data we could, for example, find the proportion of dwellings that have access to a water supply system (WSS) by department and plot it.
:::

```{r data filtering and plot, echo=TRUE, fig.width=7, fig.height=3}
# Subset data so it presents the information regarding the WSS, by department
wss <- public_services_d %>%
  subset(area == "total_departamental" & servicio_publico == "acueducto") %>%
  select(departamento, disponible, total)

# Calculate the total counts for each department (for all situations)
total_counts <- wss %>%
  group_by(departamento) %>%
  dplyr::summarise(total_all = sum(total)) %>%
  ungroup()

# Calculate the proportion of counts of "si" by department
proportions_wss <- wss %>%
  filter(disponible == "si") %>%
  left_join(total_counts, by = "departamento") %>%
  mutate(proportion_si = total / total_all)

# Change the value of cell San Andres
proportions_wss[28, "departamento"] <- "SAPSC"

# Plot proportions_wss of dwellings with access to WSS by deparment
ggplot(proportions_wss, aes(
  x = reorder(departamento, -proportion_si),
  y = proportion_si
)) +
  geom_bar(stat = "identity", fill = "#10bed2", color = "black", width = 0.6) +
  labs(
    title = "Proportion of dwellings with access to WSS, by department",
    x = "Department",
    y = "Proportion"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```

## Data Integration: Combining Datasets for Enhanced Analysis

::: {style="text-align: justify;"}
Also, with the demographic data, we could merge data from the different categories and compare information. For example, we would download a file from the persons demographic category and another one from persons social. The first one presents information regarding population's literacy and the second one regarding its marital status.
:::

```{r more data download, include=FALSE}
# Download data
literacy <- download_demographic("DANE_CNPVPD_2018_16PM")
marital_status <- download_demographic("DANE_CNPVPS_2018_7PM")
```

::: {style="text-align: justify;"}
We will focus on a random municipality, examining the total area and specifically concentrating on the age group between 15 and 24 years old. Since we want to look at both data sets at the same time, it is good to check first if they present the same count for population under the same conditions.
:::

```{r population counts, message=FALSE, warning=FALSE, results="hide"}
# Literacy data filter to have a population's count under specific conditions
lit <- literacy %>%
  subset(municipio == "Villapinzón" & grupo_de_edad == "entre_15_y_24_anos" &
    area == "total" & sabe_leer_y_escribir == "total" & sexo == "total") %>%
  select(municipio, area, grupo_de_edad, sexo, total)
head(lit)
# Marital status data filter to find population's count under certain conditions
ms <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == c("15_a_19", "20_a_24") & estado_civil == "total" &
    sexo == "total") %>%
  select(grupo_de_edad, estado_civil, sexo, total)
# Sum of counts of both age groups
ms_count <- ms %>%
  group_by(estado_civil, sexo) %>%
  dplyr::summarise(total = sum(total)) %>%
  ungroup()

# New row with summed information
new_row <- data.frame(
  grupo_de_edad = "15_a_24", estado_civil = "total",
  sexo = "total", total = toString(ms_count$total)
)
# Final dataframe with counts for each group independently and together
ms_f <- rbind(ms, new_row)
head(ms_f)
```

::: {style="text-align: justify;"}
Now that we know that both data sets contain the same counts, we can start to look at the specific information we want. For the literacy data set, we would focus on the people that do not know how to read neither write. :::
:::

```{r other data plot, echo=TRUE, fig.width=10, fig.height=3}
# Filter of literacy data to get only the counts for specific conditions
lit_filt <- literacy %>%
  subset(municipio == "Villapinzón" & sabe_leer_y_escribir == "no" &
    grupo_de_edad == "entre_15_y_24_anos" & area == "total") %>%
  select(sabe_leer_y_escribir, grupo_de_edad, sexo, total) %>%
  mutate(estado = "no_sabe_leer_y_escribir") %>%
  select(-sabe_leer_y_escribir) %>%
  mutate(proportion = total / lit$total)

# Filter of marital status data to get only the counts for specific conditions
ms_filt15 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "15_a_19") %>%
  select(grupo_de_edad, estado_civil, sexo, total)

# Filter of marital status data to get only the counts for specific conditions
ms_filt20 <- marital_status %>%
  subset(municipio == "Villapinzón" & area == "total" &
    grupo_de_edad == "20_a_24") %>%
  select(grupo_de_edad, estado_civil, sexo, total)
# Merge of both marital status dataframes
ms_filt <- rbind(ms_filt15, ms_filt20)
# Data frame with marital status info and proportions based on columns
ms_final <- ms_filt %>%
  group_by(estado_civil, sexo) %>%
  dplyr::summarise(total = sum(total)) %>%
  ungroup() %>%
  mutate(grupo_de_edad = "entre_15_y_24_anos") %>%
  rename(estado = estado_civil) %>%
  mutate(proportion = total / lit$total)

# Merged marital status and literacy final df, with subsets to a better graph
merged_df <- bind_rows(lit_filt, ms_final) %>%
  subset(sexo != "total") %>%
  subset(estado != "total") %>%
  subset(estado != "soltero_a")

colorm <- "#f04a4c"
colorh <- "#aec800"
# Plot
ggplot(merged_df, aes(x = estado, y = proportion, fill = sexo)) +
  geom_bar(stat = "identity", position = "dodge") +
  scale_fill_manual(values = c("mujer" = colorm, "hombre" = colorh)) +
  labs(
    title = "Proportion of people in Villapinzón by marital status and
    iliteracy by sex",
    x = "Status",
    y = "Count",
    fill = "Sex"
  ) +
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(hjust = 0.5)
  )
```
