---
title: "Documentation and Dictionaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Documentation and Dictionaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Understanding Datasets Names

**ColOpenData** has three  modules: demographic, geospatial and 
climate. All the available datasets have different details, aggregation and a naming system that is used to retrieve the data. All datasets within the demographic and geospatial modules need to have a source, given name, year and additional details for identification and download. They can be consulted using a common framework:

::: {.alert .alert-info}
  SOURCE_NAME_YEAR_DETAILS
:::

Currently, **ColOpenData** only includes data from official sources [(DANE)](https://www.dane.gov.co/index.php/en/) and [(IDEAM)]( http://www.ideam.gov.co/). Name, year and details are different for each 
module, and provide an unique name for each dataset.

#### Demographic

Demographic datasets are available for municipalities and departments, and contain data for Houses and Population in four categories.

- Viviendas (Dwellings)
- Hogares (Households)
- Personas Social (Persons social)
- Personas Demográfico (Persons demographic)

All datasets are retrieved from DANE, and are correspond to observed data from the National Population and Dwelling Census (CNPV). Naming is stated as follows:

- Source: **DANE**
- Name: Names include the categories
    - Viviendas: **CNPVV**
    - Hogares: **CNPVH**
    - Personas Social: **CNPVPS**
    - Personas Demográfico: **CNPVPD**
- Year: **2018**
- Details: These are related to each individual dataset, and can be related to their identifier and spatial level of aggregation. For further details please check the function `list_datasets` below.

#### Geospatial

Geospatial datasets naming is related to the level of aggregation, since they are available from Blocks to Departments. All these datasets come from DANE, and are part of the Geostatistical National Framework (MGN), which for 2018 includes a summarized version of the National Population and Dwelling Census (CNPV). Naming is stated as follows:

- Source: **DANE**
- Name: The name is related to both the MGN and the CNPV: **MGNCNPV**
- Year: **2018**
- Details: These are related to the level of aggregation:
  - Department: **DPTO**
  - Municipality: **MPIO**
  - Municipality including Class **MPIOCL**
  - Block: **MZN**
  - Rural Sector: **SETR**
  - Urban Sector: **SETU**
  - Rural Section: **SECR**
  - Urban Section: **SECU**
  - Urban Zone: **ZU**

#### Climate

This module has only one dataset, and the information required to use the related functions is related to the area of interest, dates, frequency of measurement and tags consulted. Therefore, this module does not use the same framework for naming, and the dataset name is only useful to retrieve the dictionary for the individual tags.

## List data

To check available datasets you can use the `list_datasets` function.
This function can also be filtered with the `module` parameter to indicate a specific module

```{r setup}
library(ColOpenData)
```

```{r list datasets}
# List all datasets
datasets <- list_datasets()
head(datasets)
```

```{r list demographic datasets}
# List geospatial datasets
geo_datasets <- list_datasets(module = "demographic")
head(geo_datasets)
```
Using this function we can retrieve all datasets names, source, aggregation level and small summary, to understand the contained information.

## Dictionaries

Dictionaries are provided to understand some tags and column names inside each module. These dictionaries are provided in Spanish, since they are retrieved directly from the sources.

#### Demographic

Dictionaries are not provided for this module.

#### Geospatial

Datasets inside the geospatial module contain a summarized version of the census and a dictionary is needed to understand all aggregated variables. These dictionaries contain the necessary metadata to use the available information. To retrieve them, we can use the `dictionary`function, using the dataset name as a parameter:

```{r dictionary for MGNCNPV at municipalities}
dict_mpio <- dictionary("DANE_MGNCNPV_2018_MPIO")
head(dict_mpio)
```

#### Climate

Climate data is not stored in multiple datasets but as an unique dataset with
numerous tags. These tags can also be consulted through the `dictionary`
function using the name of the climate dataset.

```{r dicionary for climate data}
dict_climate <- dictionary("IDEAM_CLIMATE_2023_MAY")
head(dict_climate)
```

Climate dictionary includes the available tags to use in the functions `download_climate`, `download_climate_geom` and `download_climate_stations`.
These tags allow the user to download specific climate attributes for the desired area of interest.
