---
title: "Documentation and Dictionaries"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{documentation_and_dictionaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```
## 1. List data
**ColOpenData** has three main  modules: demographic, geospatial and 
climate. Available datasets have different details and aggregation on each
module; however, they can be consulted using a common framework:

::: {.alert .alert-info}
  SOURCE_DATASET_YEAR_DETAILS
:::

Currently, **ColOpenData** only includes data from official sources [(DANE)](https://www.dane.gov.co/index.php/en/) and [(IDEAM)]( http://www.ideam.gov.co/). Name, year and details are different for each 
module, and provide an unique name for each dataset.

To check available datasets you can use the `list_datasets` function.
This function can also receive a string parameter to indicate a specific module

```{r setup}
library(ColOpenData)
```

```{r list datasets}
# List all datasets
list_datasets()
```

```{r list geospatial datasets}
# List geospatial datasets
list_datasets(module = "geospatial")
```

## 2. Dictionaries
### 2.1. Geospatial
Datasets inside the geospatial module contain a summarized version of the
National Population and Housing Census (CNPV) and a dictionary is needed to
understand all aggregated variables. These dictionaries contain the necessary metadata to use the available information. Year and details in geospatial datasets reference the last included national census and level of aggregation. To retrieve them, we can use the `dictionary`function, using the dataset name as a parameter:

```{r dictionary for MGNCNPV at municipalities}
dict_mpio <- dictionary("DANE_MGNCNPV_2018_MPIO")
print(dict_mpio)
```

### 2.2. Climate
Climate data is not stored in multiple datasets but as an unique dataset with
numerous tags. These tags can also be consulted through the `dictionary`
function using the name of the climate dataset. Year and details in climate datasets reference the last available backup date from source (IDEAM).

```{r dicionary for climate data}
dict_climate <- dictionary("IDEAM_CLIMATE_2023_MAY")
print(dict_climate)
```

