---
title: "Documentation and Dictionaries"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Documentation and Dictionaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Understanding Demographic and Geospatial Datasets

Demographic and geospatial data is provided from multiple datasets of the same source. To download the data, a naming framework has been implemented, which includes the source, group, year, and final details for individual identification. Details are different for every dataset and are related to the internal information they contain. The general frame can be used as follows:

::: {.alert .alert-info}
  SOURCE_GROUP_YEARS_DETAILS
:::

This naming framework is only used for the functions `download_demographic` and `download_geospatial`.

#### Demographic

Demographic datasets are available for municipalities and departments, and contain data for Dwellings and Population in four categories.

- Viviendas (Dwellings)
- Hogares (Households)
- Personas Social (Persons social)
- Personas Demográfico (Persons demographic)

All datasets are retrieved from DANE. Naming is stated as follows:

- Source: **DANE**
- Group: Names include the categories
    - Viviendas: **CNPVV**
    - Hogares: **CNPVH**
    - Personas Social: **CNPVPS**
    - Personas Demográfico: **CNPVPD**
- Year: **2018**
- Details: These are related to each individual dataset. For further details please check the function `list_datasets` below.

#### Geospatial

Geospatial datasets naming is related to the level of aggregation, since they are available from Blocks to Departments. All these datasets come from DANE, and are part of the Geostatistical National Framework (MGN), which for 2018 included a summarized version of the National Population and Dwelling Census (CNPV). Naming is stated as follows:

- Source: **DANE**
- Group: **MGN**
- Year: **2018**
- Details: These are related to the level of aggregation, and can be consulted using the function `list_datasets`.

## Understanding Climate Dataset

This module's data is stored in an unique dataset, and the information required to use the related functions is the area of interest, dates, and tags be consulted. Therefore, this module does not use the same naming framework. However, individual tags are required to download data and include:

```{r IDEAM table, echo = FALSE}
tags <- c(
  "TSSM_CON", "THSM_CON", "TMN_CON", "TMX_CON", "TSTG_CON", "HR_CAL",
  "HRHG_CON", "TV_CAL", "TPR_CAL", "PTPM_CON", "PTPG_CON", "EVTE_CON",
  "FA_CON", "NB_CON", "RCAM_CON", "BSHG_CON", "VVAG_CON", "DVAG_CON",
  "VVMXAG_CON", "DVMXAG_CON"
)
variable <- c(
  "Dry-bulb Temperature", "Wet-bulb Temperature",
  "Minimum Temperature", "Maximum Temperature",
  "Dry-bulb Temperature (Termograph)", "Relative Humidity",
  "Relative Humidity (Hydrograph)", "Vapour Pressure", "Dew Point",
  "Precipitation (Daily)", "Precipitation (Hourly)", "Evaporation",
  "Atmospheric Phenomenon", "Cloudiness", "Wind Trajectory",
  "Sunshine Duration", "Wind Speed", "Wind Direction",
  "Maximum Wind Speed", "Maximum Wind Direction"
)

IDEAM_tags <- data.frame(
  Tags = tags, Variable = variable,
  stringsAsFactors = FALSE
)
knitr::kable(IDEAM_tags)
```

These tags are meant to be used for download using `download_climate`, `download_climate_geom` and `download_climate_stations`.

## List data

To check available datasets you can use the `list_datasets` function.
This function can also be filtered with the `module` parameter to indicate a specific module

```{r setup}
library(ColOpenData)
```

```{r list datasets}
# List all datasets
datasets <- list_datasets()
head(datasets)
```

```{r list demographic datasets}
# List only geospatial datasets
geo_datasets <- list_datasets(module = "demographic")
head(geo_datasets)
```

Using this function, we can retrieve all names, source, aggregation level and information for individual datasets.

## Dictionaries

Dictionaries are provided to understand some tags and column names inside each module. These dictionaries are provided in Spanish, since they are retrieved directly from the sources.

#### Demographic

Dictionaries are not provided for this module.

#### Geospatial

Datasets inside the geospatial module contain a summarized version of the census and a dictionary is needed to understand all aggregated variables. These dictionaries contain the necessary metadata to use the available information. To retrieve them, we can use the `dictionary`function, using the dataset name as a parameter:

```{r dictionary for MGNCNPV at municipalities}
dict_mpio <- dictionary("DANE_MGN_2018_MPIO")
head(dict_mpio)
```

#### Climate

Climate data is not stored in multiple datasets but as an unique dataset with
numerous tags. These tags can also be consulted through the `dictionary`
function using the name of the only climate dataset.

```{r dicionary for climate data}
dict_climate <- dictionary("IDEAM_CLIMATE_2023_MAY")
head(dict_climate)
```

