---
title: "Maps and plots with ColOpenData"
output: 
  rmarkdown::html_vignette:
    df_print: tibble
vignette: >
  %\VignetteIndexEntry{Maps and plots with ColOpenData}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

::: {style="text-align: justify;"}
**ColOpenData** can be used to access open geospatial data from Colombia. This data is retrieved from the Geostatistical National Framework (MGN), published by the National Administrative Department of Statistics [(DANE)](https://www.dane.gov.co/index.php/en/). The MGN contains the political-administrative division and is used to reference census statistical information. Further information can be obtained directly from DANE [here](https://www.dane.gov.co/index.php/actualidad-dane/5454-el-dane-actualizo-el-marco-geoestadistico-nacional-a-2018).

This package contains the 2018's version of the MGN, which also included a summarized version of the National Population and Dwelling Census (CNPV) in different aggregation levels. Each level is stored in a different dataset, which can be retrieved using the `download_geospatial` function, which requires three arguments:

- `dataset` character with the geospatial dataset name.
- `include_geom` logical for including (or not) geometry. Default is `TRUE`.
- `include_cnpv` logical for including (or not) CNPV demographic and socioeconomic information. Default is `TRUE`.

To better understand dataset names and details go to [Documentation and Dictionaries](https://epiverse-trace.github.io/ColOpenData/articles/documentation_and_dictionaries.html).

Details for geospatial datasets relate to the level of aggregation as follows:
:::

```{r available datasets, echo = FALSE}
code <- c(
  "DANE_MGN_2018_DPTO", "DANE_MGN_2018_MPIO",
  "DANE_MGN_2018_MPIOCL", "DANE_MGN_2018_MZN",
  "DANE_MGN_2018_SECR", "DANE_MGN_2018_SECU",
  "DANE_MGN_2018_SETR", "DANE_MGN_2018_SETU",
  "DANE_MGN_2018_ZU"
)
level <- c(
  "Department", "Municipality",
  "Municipality including Class", "Block", "Rural Sector", "Urban Sector",
  "Rural Section", "Urban Section", "Urban Zone"
)

mgncnpv <- data.frame(
  Code = code, Level = level,
  stringsAsFactors = FALSE
)
knitr::kable(mgncnpv)
```

::: {style="text-align: justify;"}
In this vignette you will learn:
1. How to download geospatial data using **ColOpenData**
2. How to use census data included in geospatial datasets
3. How to visualize spatial data using **leaflet** and **ggplot2**

We will be using geospatial data at the level of Urban Section (SECU) for the municipality of Tunja and will calculate the percentage of houses with internet connection at each SECU. Later, we will build some plots using the previously mentioned approaches for dynamic and static plots.

We will start by importing the needed libraries.

```{r library imports, results = "hide", warning = FALSE, message = FALSE}
library(ColOpenData)
library(dplyr)
library(sf)
library(ggplot2)
library(leaflet)
```

::: {style="text-align: justify;"}
**Disclaimer: all data is loaded to the environment in the user's R session, but is not downloaded to user's computer.**
:::

## Downloading geospatial data

First, we download the data using the function `download_geospatial`, including the geometries and the census related information.
:::

```{r download data}
# Download complete dataset
secu <- download_geospatial(
  dataset = "DANE_MGN_2018_SECU",
  include_geom = TRUE,
  include_cnpv = TRUE
)
head(secu)
```

::: {style="text-align: justify;"}
After downloading, we have to filter by the municipality code using the DIVIPOLA code for Tunja. DIVIPOLA codification is standardized for the whole country, and contains department's and municipalities' codes. Departments have two digits for individual identification, while municipalities have five. The codes for each municipality and department can be consulted in the following [table](https://www.datos.gov.co/widgets/gdxc-w37w). 

To get the DIVIPOLA code of a municipality or department we can use the auxiliary functions `divipola_municipality_code` and `divipola_department_code` in **ColOpenData**. To retrieve a municipality code we must include the department name and the municipality name. This is to consider repetition in municipalities' names across departments.

```{r}
divipola_municipality_code("BOYACA", "TUNJA")
```

We can see that the DIVIPOLA code for Tunja is `"15001"`. We can use this code for filtering.

To understand which column contains the municipalities' codes and filter for Tunja, we will need the corresponding dataset dictionary. To download the dictionary we can use the `dictionary` function. This function uses the dataset name to download the associated information. For further information please refer to the documentation on dictionaries previously mentioned.
:::

```{r dictionary for urban sections}
# Urban section dictionary
dict <- dictionary("DANE_MGN_2018_SECU")
head(dict)
```

::: {style="text-align: justify;"}
After exploring the dictionary, we can identify the column that contains the individual municipality codes is [MPIO_CDPMP]{.underline}. This column contains the concatenated code for department and municipality, which is a five digit unique identifier for every municipality in the country.

To calculate the percentage of houses with internet connection, we will need to know the number of houses with internet connection and the total of houses in each SECU. From the dictionary we get that the number of houses with internet connection is [STP19_INT1]{.underline} and the total of houses is [STVIVIENDA]{.underline}. We will calculate the percentage as follows:
:::

```{r filter and internet coverage}
# Filter for Tunja
tunja <- secu %>% filter(MPIO_CDPMP == "15001")

# Calculate internet coverage percentage
tunja <- tunja %>% mutate(INT_PERC = round(STP19_INT1 / STVIVIENDA, 2))
```

## Static plots (ggplot2)

::: {style="text-align: justify;"}
[`ggplot2`](https://ggplot2.tidyverse.org/) can be used to generate static plots of spatial data by using the geometry `geom_sf` as follows:

```{r ggplot2 basic}
ggplot(data = tunja) +
  geom_sf(mapping = aes(fill = INT_PERC), color = NA)
```

Color palettes and themes can be defined for each plot using the aesthetic and scales, which can be consulted in the `ggplot2` [documentation](https://ggplot2.tidyverse.org/reference/index.html#scales).
:::

```{r ggplot2}
# ggplot2
ggplot(data = tunja) +
  geom_sf(mapping = aes(fill = INT_PERC), color = NA) +
  theme_minimal() +
  theme(
    panel.grid = element_blank(),
    axis.text = element_blank(),
    axis.ticks = element_blank()
  ) +
  scale_fill_gradient("Percentage", low = "#10bed2", high = "#deff00") +
  ggtitle(
    label = "Internet coverage",
    subtitle = "Tunja, Colombia"
  )
```

## Dynamic plots (leaflet)

::: {style="text-align: justify;"}
For dynamic plots, we can use [`leaflet`](https://rstudio.github.io/leaflet/), which is an open-source library for interactive maps. To create the same plot we first will create a color palette.
:::

```{r define color palette}
# Color Palette
colfunc <- colorRampPalette(c("#10bed2", "#deff00"))
pal <- colorNumeric(
  palette = colfunc(100),
  domain = tunja$INT_PERC
)
```

::: {style="text-align: justify;"}
With the previous color palette we can generate the interactive plot. The package also includes open source maps for the base map like [OpenStreetMap](https://www.openstreetmap.org/#map=5/4.632/-74.299) and [CartoDB](https://carto.com/basemaps). For further details on `leaflet`, please refer to the package's [documentation](https://rstudio.github.io/leaflet/reference/index.html).
:::

```{r leaflet}
# Leaflet
leaflet(tunja) %>%
  addProviderTiles(providers$CartoDB.Positron) %>%
  addPolygons(
    stroke = TRUE,
    weight = 0,
    color = NA,
    fillColor = ~ pal(tunja$INT_PERC),
    fillOpacity = 1,
    popup = paste0(tunja$INT_PERC)
  ) %>%
  addLegend(
    position = "bottomright",
    pal = pal,
    values = ~ tunja$INT_PERC,
    opacity = 1,
    title = "Internet Coverage"
  )
```
