---
title: "Download geospatial data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Download geospatial data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

**ColOpenData** can be used to access open Colombian geospatial data. This 
data is retrieved from the Geostatistical National Framework (MGN), published by
the Administrative National Department of Statistics
[(DANE)](https://www.dane.gov.co/index.php/en/). The MGN contains the political-administrative division and is used to reference census statistical information. Further information can be obtained directly from DANE [here](https://www.dane.gov.co/index.php/actualidad-dane/5454-el-dane-actualizo-el-marco-geoestadistico-nacional-a-2018)

This package contains the 2018's version of the MGN, which also included a
summarized version of the National Population and Housing Census (CNPV) in
different aggregation levels. Each level is stored in a different dataset, 
which can be retrieved using the `download_geospatial` function.

`download_geospatial` only requires one argument, which is the name of the
dataset to consult. To better understand dataset names and metadata go to the
[Documentation and Dictionaries](https://epiverse-trace.github.io/ColOpenData/articles/documentation_and_dictionaries.html).
Details for geospatial datasets relate to the level of aggregation as follows:

```{r Available datasets, echo = FALSE}
code <- c(
  "DANE_MGNCNPV_2018_DPTO", "DANE_MGNCNPV_2018_MPIO",
  "DANE_MGNCNPV_2018_MPIOCL", "DANE_MGNCNPV_2018_MZN",
  "DANE_MGNCNPV_2018_SECR", "DANE_MGNCNPV_2018_SECU",
  "DANE_MGNCNPV_2018_SETR", "DANE_MGNCNPV_2018_SETU"
)
level <- c(
  "Department", "Municipality",
  "Municipality including Class", "Block Level", "Rural Sector", "Urban Sector",
  "Rural Section", "Urban Section"
)

mgncnpv <- data.frame(
  Code = code, Level = level,
  stringsAsFactors = FALSE
)
knitr::kable(mgncnpv)
```

For this example, we will download the geospatial data at the level of
department, to later include epidemiological data from dengue cases in 2018
and visualize the departments with the highest case counts.

We will start by importing needed libraries
```{r library imports, results = "hide", warning = FALSE, message = FALSE}
library(ColOpenData)
library(dplyr)
library(sf)
library(ggplot2)
library(RColorBrewer)
```


## Downloading and plotting geospatial data

To download the geospatial data we will use the `download_geospatial` function,
with the respective code for department level.
```{r download mpios, results = "hide", warning = FALSE, message = FALSE}
departments <- download_geospatial("DANE_MGNCNPV_2018_DPTO")
```
```{r}
str(departments)
```
Once downloaded, we can subset the data as desired. In this example we only 
need the departments codes and geometries.

```{r}
departments_filtered <- departments %>%
  select(DPTO_CCDGO, geometry)
```
```{r}
str(departments_filtered)
```


Dengue cases during 2018 can be loaded from the example dataset `dengue_2018`,
which was previously loaded in the package.

```{r}
data(dengue_2018)
```
```{r}
str(dengue_2018)
```

To merge the datasets we will use the department’s codes, marked by the same 
column name in this example.

```{r}
dengue_col <- merge(
  x = departments_filtered,
  y = dengue_2018,
  by = "DPTO_CCDGO"
)
```

Finally, we can visualize the cases in a map.
```{r}
ggplot(data = dengue_col) +
  geom_sf(mapping = aes(fill = CASES)) +
  scale_fill_gradientn(
    colours = brewer.pal(7, "YlOrRd"),
    name = "Cases"
  ) +
  ggtitle("Dengue Cases in Colombia during 2018") +
  theme_bw()
```

